---
title: "STA 631 Statistical Modelling And Regression Portfolio"
site: distill::distill_website
---

```{r setup }
knitr::opts_chunk$set(echo = TRUE)

```

### Loading the libraries 

```{r}
# load the libraries
library(tidyverse)
library(dplyr)
library(corrplot)
library(caret)
library(Metrics)
library(rsq)
library(pROC)
```

### Load the Dataset
```{r }
# load the dataset
hotel_data <- read_csv("hotel_bookings.csv")
```
### Summary of the dataset 
```{r}
# overview of the dataset 

glimpse(hotel_data)

```

#### Removing missing values columns from the dataset 

```{r}
null_counts <- colSums(is.na(hotel_data))
null_counts
```


Here we can see the in children column 4 observations are missing, so we will remove the null value rows to cleanup the dataset

```{r}
hotel_clean_data <- na.omit(hotel_data)

skimr::skim(hotel_clean_data)

```
### Data Visualization 

#### 1) What is the range of years covered by the dataset, including both the earliest and latest years?

```{r}
# let's find the years the data taken from 
years_of_data <- unique(hotel_clean_data$arrival_date_year)
print(years_of_data)

```

Ans) The data is taken from 2015, 2016, 2017 approximately more than 1 lakh columns are recorded 


#### 2) What is the total number of bookings canceled and the total number of bookings not canceled? 

```{r}
# group by the  booking canceleation status 
grp_by_canceled_hotel <- hotel_clean_data %>%
  group_by(is_canceled) %>%
  summarise(count = n())

grp_by_canceled_hotel

```

Ans) From the above table we can see that there are 75166  hotel bookings are not canceled  and 44220 hotel bookings are canceled

#### 3) what percentage of bookings are canceled in the total bookings?

```{r}
# Calculate the total number of bookings
total_bookings <- sum(grp_by_canceled_hotel$count)

# Calculate the number of canceled bookings
canceled_bookings <- grp_by_canceled_hotel$count[grp_by_canceled_hotel$is_canceled == 1]

# Calculate the percentage of bookings that are canceled overall
percentage_canceled <- (sum(canceled_bookings) / total_bookings) * 100

# Print the result
cat("The percentage of bookings that are canceled overall is:", percentage_canceled, "%\n")

```

#### 4) Distribution of Canceled and Not Canceled Bookings

```{r}

# Calculate the total number of bookings
total_bookings <- sum(grp_by_canceled_hotel$count)

# Calculate the percentage of each status type
percentage <- round((grp_by_canceled_hotel$count / total_bookings) * 100, 2)

# Create labels with both status type and percentage
labels <- paste(c( "Not Canceled","Canceled"), "(", percentage, "%)", sep = " ")

# Define colors for each slice
colors <- c("green", "red")  # You can customize the colors as needed

# Plot the pie chart with labels and colors
pie(grp_by_canceled_hotel$count, labels = labels, col = colors,
    main = "Distribution of Canceled and Not Canceled Bookings")


```

#### 5) Total Number of Bookings Canceled vs. Not Canceled

```{r}
# Create a ggplot object
ggplot(hotel_clean_data, aes(x = factor(is_canceled), fill = factor(is_canceled))) +
  geom_bar() +
  labs(title = "Total Number of Bookings Canceled vs. Not Canceled",
       x = "Cancellation Status",
       y = "Total Number of Bookings") +
  scale_x_discrete(labels = c("Not Canceled", "Canceled"))

```

#### 6) What are the combined totals of bookings that were canceled and not canceled, categorized by hotel type?

```{r}
# group by the hotel type and cancelation status  
grp_by_hotel_type <- hotel_clean_data %>%
  group_by(hotel,is_canceled) %>%
  summarise(count = n())

grp_by_hotel_type
```

#### 7) Total Number of Bookings by Cancellation Status and Hotel Type

```{r}
# Create a ggplot object
ggplot(hotel_clean_data, aes(x = factor(is_canceled), fill = factor(hotel))) +
  geom_bar(position = "dodge") +
  labs(title = "Total Number of Bookings by Cancellation Status and Hotel Type",
       x = "Cancellation Status",
       y = "Total Number of Bookings") +
  scale_x_discrete(labels = c("Not Canceled", "Canceled"))


```

#### 8) What are the combined totals of bookings that were canceled and not canceled, categorized by year ?

```{r}
# group by the year and cancelation status 
grp_by_year <- hotel_clean_data %>%
  group_by(arrival_date_year,is_canceled) %>%
  summarise(count = n())

grp_by_year
```

#### 9) Total Number of Bookings by Cancellation Status over Years

```{r}
# Filter for canceled and not canceled bookings separately
canceled_bookings <- grp_by_year %>% filter(is_canceled == 1)
not_canceled_bookings <- grp_by_year %>% filter(is_canceled == 0)

# Create line plot
ggplot() +
  geom_line(data = canceled_bookings, aes(x = arrival_date_year, y = count, color = "Canceled"), size = 1) +
  geom_line(data = not_canceled_bookings, aes(x = arrival_date_year, y = count, color = "Not Canceled"), size = 1) +
  labs(
    title = "Total Number of Bookings by Cancellation Status over Years",
    x = "Year",
    y = "Total Number of Bookings") +
  scale_color_manual(values = c("red", "blue"), name = "Cancellation Status",
                     labels = c("Canceled", "Not Canceled")) +
  theme_minimal()

```

From the above graph we can see that from 2015 to 2016 cancelations and not cancelations are increased and from 2016 to 2017 cancelation and not cancelations are decreased.

#### 10) What percentage of bookings canceled in each month ?

```{r}
# Group data by 'arrival_date_month' and 'is_canceled', then calculate the total number of bookings
total_bookings <- hotel_clean_data %>%
  group_by(arrival_date_month) %>%
  summarise(total = n())

# Group data by 'arrival_date_month' and 'is_canceled', then calculate the number of canceled bookings
canceled_bookings <- hotel_clean_data %>%
  filter(is_canceled == 1) %>%
  group_by(arrival_date_month) %>%
  summarise(canceled_total = n())

# Merge total and canceled bookings dataframes
merged_data <- merge(total_bookings, canceled_bookings, by = "arrival_date_month", all.x = TRUE)

# Calculate the percentage of canceled bookings
merged_data$percentage_canceled <- (merged_data$canceled_total / merged_data$total) * 100

# Print the resulting dataframe
print(merged_data)
```

#### 11) Percentage of Bookings Canceled over Month 

```{r}

# Convert 'arrival_date_month' to factor with correct order
merged_data$arrival_date_month <- factor(merged_data$arrival_date_month, 
                                         levels = c("January", "February", "March", "April",  "May","June", "July", "August", "September", "October", "November", "December"))

# Create a ggplot object
ggplot(merged_data, aes(x = arrival_date_month, y = percentage_canceled, group = 1)) +
  geom_line(color = "blue") +
  geom_point(color = "blue", size = 3) +
  labs(
    title = "Percentage of Bookings Canceled over Month ",
    x = "Month",
    y = "Percentage Canceled") +
  scale_y_continuous(labels = scales::percent_format(scale = 1, accuracy = 1)) +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1))



```


#### 12) What are the types of rooms that were booked by customers and subsequently canceled? 

```{r}
grp_by_roomtype <- hotel_clean_data %>%
  group_by(reserved_room_type,is_canceled) %>%
  summarise(count = n())

grp_by_roomtype
```

#### 13) Number of Bookings by Reserved Room Type and Cancellation Status

```{r}
ggplot(grp_by_roomtype, aes(x = reserved_room_type, y = count, fill = factor(is_canceled))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Bookings by Reserved Room Type and Cancellation Status",
       x = "Reserved Room Type",
       y = "Number of Bookings",
       fill = "Cancellation Status") +
  scale_fill_manual(values = c("Not Canceled" = "skyblue", "Canceled" = "salmon")) +
  facet_grid(. ~ factor(is_canceled)) +  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

#### Dropping Unwanted columns

```{r}
df <- hotel_clean_data %>%
  select(
    -arrival_date_year,
    -arrival_date_month,
    -children,
    -babies,
    -meal,
    -country,
    -market_segment,
    -distribution_channel,
    -agent,
    -company,
    -reservation_status,
    -required_car_parking_spaces,
    -reservation_status_date,
    -assigned_room_type
  )
```

### Correlation Matrix

```{r}
# Filter numeric columns
df_numeric <- df[, sapply(df, is.numeric)]

# Compute correlation matrix
correlation_matrix <- cor(df_numeric, use = "pairwise.complete.obs")

print(correlation_matrix)


```

### HeatMap 

```{r}
# Plot heatmap
corrplot(correlation_matrix, method = "color", type = "upper", 
         tl.col = "black", tl.srt = 45, tl.cex = 0.7, addCoef.col = "black", 
         number.cex = 0.7, col = colorRampPalette(c("white", "red"))(100),
         title = "Correlation Heatmap")
```



### Splitting the Dataset As Test Dataset  and Train Dataset

```{r}


# Set seed for reproducibility
set.seed(100)

# Create data partition
test_index <- createDataPartition(y = df$is_canceled, times = 1, p = 0.3, list = FALSE)

# Create training dataset
df_train <- df[-test_index,]

# Create validation dataset
df_validation <- df[test_index,]

# Check dimensions of training dataset
dim(df_train)
```

### Multiple linear Regression Model 

```{r}
# Train multiple-linear regression model
mlr_model <- lm(is_canceled ~ ., data = df_train)

# Print summary of the model
summary(mlr_model)

```
```{r}
# Make predictions on the validation dataset
mlr_pred <- predict(mlr_model, newdata = df_validation, type = "response")

# Convert predicted probabilities to binary predictions
mlr_pred <- ifelse(mlr_pred > 0.5, 1, 0)

table(df_validation$is_canceled==mlr_pred)

# Print accuracy
accuracy <- sum(mlr_pred == df_validation$is_canceled) / nrow(df_validation)
print(paste("Accuracy:\n", accuracy))

conf_matrix <- confusionMatrix(factor(mlr_pred), factor(df_validation$is_canceled))

# Print confusion matrix
print(conf_matrix)


```

##### Performance Accuracy of the Multi Linear Regression Model is 76 %

### Logistic Regression Model

```{r}

# Train logistic regression model
lr_model <- glm(is_canceled ~ ., data = df_train, family = binomial)

summary (lr_model)

```



```{r}
# Make predictions on the validation dataset
lr_pred <- predict(lr_model, newdata = df_validation, type = "response")

# Convert predicted probabilities to binary predictions
lr_pred <- ifelse(lr_pred > 0.5, 1, 0)

table(df_validation$is_canceled==lr_pred)

# Print accuracy
accuracy <- sum(lr_pred == df_validation$is_canceled) / nrow(df_validation)
print(paste("Accuracy:\n", accuracy))

conf_matrix <- confusionMatrix(factor(lr_pred), factor(df_validation$is_canceled))

# Print confusion matrix
print(conf_matrix)


```

##### Performance Accuracy of the Logistic Model is 78 %


```{r}
# Calculate confusion matrix values
TN <- conf_matrix$table[1, 1]  # True negatives
FP <- conf_matrix$table[1, 2]  # False positives
FN <- conf_matrix$table[2, 1]  # False negatives
TP <- conf_matrix$table[2, 2]  # True positives

# Calculate precision
precision <- TP / (TP + FP)

# Calculate recall
recall <- TP / (TP + FN)

# Calculate F1 score
f1_score <- 2 * (precision * recall) / (precision + recall)

# Print F1 score
print(paste("F1 Score:\n", f1_score))

```


### RoC Curve

```{r}



# Create ROC curve object
lr_roc <- roc(df_validation$is_canceled, lr_pred)

# Plot ROC curve
plot(lr_roc, print.auc = TRUE, print.auc.y = 0.4, col = "blue")

# Add legend
legend("bottom", legend = c("Logistic Regression"), col = c("blue"), lwd = 2, horiz = TRUE)

```
An AUC (Area Under the Curve) value of 0.723 indicates the performance of the logistic regression model in distinguishing between the positive class hotel booking cancellations)and the negative class no cancellations based on the predicted probabilities

in summary an AUC of 0.723 indicates that the logistic regression model shows moderate discriminative power in predicting hotel booking cancellations. While the model performs better, there may be opportunities to enhance its predictive ability further

### Course Objectives 

•	Describe probability as a foundation of statistical modeling, including inference and maximum likelihood estimation

•	Determine and apply the appropriate generalized linear model for a specific data context

•	Conduct model selection for a set of candidate models

•	Communicate the results of statistical models to a general audience

•	Use programming software (i.e., R) to fit and assess statistical models



#### 1.	Describe probability as a foundation of statistical modeling, including inference and maximum likelihood estimation:

Probability theory serves as a fundamental idea in statistical modeling for both logistic regression and multiple linear regression. In logistic regression, the logistic function is used to explicitly model probabilities, especially for binary outcomes. Probabilities play an indirect role in multiple linear regression (MLR) by helping to estimate coefficients and make predictions based on linear connections between variables.

#### 2.	Determine and apply the appropriate generalized linear model for a specific data context:

Logistic regression is best suited for binary classification tasks like predicting hotel booking cancellations. It's a type of generalized linear model (GLM) designed for binary outcomes.

#### 3.	Conduct model selection for a set of candidate models:

Model selection involves comparing the performance of different models. In this context, I have  
compared the performance of the MLR and logistic regression models using evaluation metrics such as accuracy, precision, recall, F1-score, The model with the highest performance metrics is selected as the best model for predicting hotel booking cancellations.
	

#### 4.	Communicate the results of statistical models to a general audience:

Our analysis says that there are many factors are involved booking cancelation, like lead time, previous cancelation, and waiting list etc.

•	Interpretation of Coefficients:
MLR: Each coefficient represents the change in the dependent variable (is_canceled) associated with a one-unit change in the predictor variable, holding all other variables constant. For example, a one-unit increase in lead_time is associated with a 0.0006431 increase in the log-odds of cancellation.
	
Logistic Regression: Coefficients represent the log-odds of the dependent variable (is_canceled) being 1 (cancellation) compared to 0 (no cancellation). For instance, a one-unit increase in lead_time is associated with a 0.0038772 increase in the log-odds of cancellation.
	
•	Model Performance:
Accuracy: The logistic regression model (78.49%) outperforms the MLR model (76.26%) in terms of overall accuracy.
F1 Score: The F1 score, a measure of a model's accuracy, considers both the precision and recall of the model. The logistic regression model has an F1 score of 0.624, indicating better balance between precision and recall compared to the MLR model.
	
•	Significance of Predictors:
Both models indicate significant predictors for hotel booking cancellations, such as lead_time, previous_cancellations, deposit_type, and customer_typeTransient. However, the logistic regression model provides odds ratios, which can be directly interpreted as the change in odds of cancellation associated with a one-unit change in the predictor.

#### 5.	Use programming software (i.e., R) to fit and assess statistical models:

Both logistic regression and MLR models are implemented using programming software such as R. i have used R to fit these models to your data, assess model assumptions, evaluate model performance, and interpret results. Through hands-on experience with R, i have gained proficiency in using programming software to analyze and assess statistical both models.





